---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

LINCS analysis Vignette.

In order to complete this analysis, the level 4 dataset should be obtained from the LINCS project:

https://docs.google.com/forms/d/1j6Vb_s4FrDodxoS3IDZsHWoVNrOnKNQbqwbwcW2a208/viewform

The expression data matrix is large (~120GB), and may be obtained using the s3cmd command line tool on Linux-like systems:

# configure s3cmd with your keys
s3cmd --configure

# now get the data file.  takes 30-60 minutes with a fast data connection.
s3cmd get  s3://data.lincscloud.org/l1000/level4/zspc_n1328098x22268.gctx
s3cmd get  s3://data.lincscloud.org/l1000/metadata/inst.info

The ExpSpect package requires the following packages:

install.packages(R6, rhdf5, hgu133plus2.db, RCurl, RJSONIO)

The instance metadata from LINCS is provided as tab delimited file (inst.info) which is unwieldy to work with in an R session due to it size.  Therefore, the first step is to convert this file to hdf5 format using a helper function provided by the ExpSpect package:

lincs <- LINCS$new("/path/to/zspc_n1328098x22268.gctx")
lincs$info2hdf5("/path/to/inst.info", "/desired/path/to/inst_info.gctx")
# now tell your lincs object where you saved the file just created
lincs$setInfoFile("/path/to/inst_info.gctx")

Now we are ready to extract some signatures of interest from the LINCS expression data.  First we will restrict our analysis to normal cell lines.  

normcells <- lincs$cellInfo() # fetches normals by default

The ExpSpect package provides a function to retrieve meta data from the info file.  However, for large contiguous reads (in this case, all columns for 1 row), it is much much faster to use th hdf5 call directly:

cellids <- h5read(lincs$infoFile, "0/DATA/0/matrix", start=c(4,1), count=c(1,lincs$ncol))
ixnorm <- which(cellids[1,] %in% normcells$cellid)

Next let us restrict the analysis to instances where the perturbagen was a clinically relevant drug.  This is a fuzzy concept but can be approximated by searching for those perturbagens with pert_type of trt_cp (chemical compound) and has a common name as opposed to an alphanumeric code (i.e. pert_iname is comprised of lower case characters).

fdadrugs <- lincs$fdaDrugs()
pertids <- h5read(lincs$infoFile, "0/DATA/0/matrix", start=c(2,1), count=c(1,lincs$ncol))[ixnorm]
ixfda <- which(pertids %in% fdadrugs$pertid)

Now we can extract the datamatrix for these instances.

lincs$getData()


