---
title: "ExpSpect Analysis Example"
author: "Eric J. Kort"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ExpSpect Analysis Example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Analysis of LINCS data with the ExpSpect package

### Preparing your environment

In order to complete this analysis, the level 4 dataset should be obtained from the [lincs project](https://docs.google.com/forms/d/1j6Vb_s4FrDodxoS3IDZsHWoVNrOnKNQbqwbwcW2a208/viewform).  In addition, the ExpSpect package requires the following packages: `R6`, `rhdf5`, `hgu133plus2.db`, `RCurl`, `RJSONIO`.

The expression data matrix is large (~120GB), and may be obtained using the s3cmd command line tool on Linux-like systems:

```
# configure s3cmd with your keys
s3cmd --configure

# now get the data file.  takes 30-60 minutes with a fast data connection.
s3cmd get  s3://data.lincscloud.org/l1000/level4/zspc_n1328098x22268.gctx
s3cmd get  s3://data.lincscloud.org/l1000/metadata/inst.info
```
The instance metadata from LINCS is provided as tab delimited file (inst.info) which is unwieldy to work with in an R session due to it size.  Therefore, the first step is to convert this file to hdf5 format using a helper function provided by the ExpSpect package:

```{r, eval=TRUE}
library(ExpSpect)
# adjust path below as needed.
lincs <- LINCS$new("/mnt/lincs/zspc_n1328098x22268.gctx")

## perform the following after downloading the data.  Only needs to be done once.
#lincs$info2hdf5("/mnt/lincs/inst.info", "/desired/path/to/inst_info.gctx")

# now tell your lincs object where you saved the file just created
lincs$setInfoFile("/mnt/lincs/inst_info.gctx")
```

Now we are ready to begin.

### Feature reduction

First we will extract some signatures of interest from the LINCS expression data.  First we will restrict our analysis to normal cell lines.  

```{r}
normcells <- lincs$cellInfo() # fetches normals by default
```

The ExpSpect package provides a function to retrieve meta data from the info file.  However, for large contiguous reads (in this case, all columns for 1 row), it is much much faster to use th hdf5 call directly:

```{r}
cellids <- h5read(lincs$infoFile, "0/DATA/0/matrix", start=c(4,1), count=c(1,lincs$ncol))
ixnorm <- which(cellids[1,] %in% normcells$cellid)
```

Next let us restrict the analysis to instances where the perturbagen was a clinically relevant drug.  Again, the `ExpSpect` package contains a utility function that identifies drugs among the LINCS perturbagens that are FDA approved for one or more indications.  Here we will identify the relevant `pertids` for the normal samples from above and extract the corresponding expression data.

```{r}
fdadrugs <- lincs$fdaDrugs()
pertids <- h5read(lincs$infoFile, "0/DATA/0/matrix", start=c(2,1), count=c(1,lincs$ncol))[ixnorm]
ixfda <- which(pertids %in% fdadrugs$pertid)
datamatrix <- lincs$getData(1:100, ixnorm[ixfda])
```

## No User Serviceable Parts Below This Line

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.


Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))